---
title: "R Notebook"
output: html_notebook
---

Component 3

steps:importing data
      removing null values
      removing outliers
      scaling the data to normalize the data
      Verifying all data points in each field are of same format

```{r}
library(dplyr)
library(ggplot2)
library(readxl)
library(lubridate)
library(caret)
setwd("G:/LILTHOMA/Rise_Wpu/Mentor_Mind/Identify_customer_segments_for_online_retail_using_K-means")
customer_data <- read_excel("Online Retail.xlsx")
View(customer_data)

null_values <- colSums(is.na(customer_data))
```


```{r}
null_df <- data.frame(null_values)
print(sum(duplicated(customer_data)))
null_df%>%mutate(percentage=null_values/nrow(customer_data))

```
 #CustomerID has 24% data missing and Description has less than 1% missing values
 #5268 duplicated values are present
```{r}
clean_df <- customer_data%>%
  filter_all(all_vars(!is.na(.)))
removed_data <- customer_data%>%
  filter_all(any_vars(is.na(.)))
clean_df
```
Now we check the str of data to understand the format of each field if there is any changes required we will implement the and edit where changes are required



```{r}
#Removing Duplicates
final_df <-distinct(clean_df)
print(sum(duplicated(final_df)))
```
#We have data from december 2010 to november 2011
```{r}
final_df$month <-month(final_df$InvoiceDate)
summary(final_df$month)
```

```{r}
#finding outliers UnitPrice

q1_price <- quantile(final_df$UnitPrice,0.25,na.rm=T)
q3_price <-quantile(final_df$UnitPrice,0.75,na.rm=T)
iqr<-q3_price - q1_price
lower_bound_price = q1_price - 1.5*iqr
upper_bound_price =q3_price + 1.5*iqr
outliers <- final_df$UnitPrice[final_df$UnitPrice < lower_bound_price | final_df$UnitPrice > upper_bound_price]
final_df$UnitPrice <- ifelse(final_df$UnitPrice %in% outliers, NA, final_df$UnitPrice)


```


```{r}
boxplot(final_df$UnitPrice)
```


```{r}
#finding outliers Quantity

q1_quant <- quantile(final_df$Quantity,0.25,na.rm=T)
q3_quant <-quantile(final_df$Quantity,0.75,na.rm=T)
iqr<-q3_quant - q1_quant
lower_bound_quant = q1_quant - 1.5*iqr
upper_bound_quant =q3_quant + 1.5*iqr
outliers_quant <- final_df$Quantity[final_df$Quantity < lower_bound_quant | final_df$Quantity > upper_bound_quant]
final_df$Quantity <- ifelse(final_df$Quantity %in% outliers_quant, NA, final_df$Quantity)
boxplot(final_df$Quantity)

```


```{r}
sum(is.na(final_df))
new_data <- final_df%>%filter_all(all_vars(!is.na(.)))
sum(is.na(new_data))
```
```{r}
print(new_data%>%group_by(CustomerID)%>%count()%>%arrange(-n)%>%head(30))
new_data%>%group_by(StockCode)%>%count()%>%arrange(-n)%>%head(30)
new_data%>%group_by(InvoiceNo)%>%count()%>%arrange(-n)%>%head(30)
sum(duplicated(new_data))

```


#We have finished the data preperation process
1)removed null values
2)extracted month values
3)removed outiers from the data

#doubts 
invoice number stock code and customer Id are the unique fields

```{r}
library(stringr)
new_data$StockCode <- str_replace(new_data$StockCode, "[A-Za-z]$", "")
new_data$StockCode <- as.numeric(new_data$StockCode)
new_data$InvoiceNo <- str_replace(new_data$InvoiceNo, "[A-Za-z]$", "")
new_data$InvoiceNo <- as.numeric(new_data$InvoiceNo)
str(new_data)


#Final cleaned values with no inconsistent data formats and no null or dupicate values
cleaned_data <-new_data%>%filter_all(all_vars(!is.na(.)))
sum(is.na(cleaned_data))

```

```{r}
str(cleaned_data)
selected_df <- cleaned_data%>%select(c(InvoiceNo,StockCode,month,UnitPrice,Quantity))
selected_df_scaled <-scale(selected_df,center=TRUE)

```



```{r}
#install.packages("factoextra")
library(factoextra)
#methods : wss silhouette gap_stat

```
```{r}

find_k <-selected_df 
set.seed(123)
sample_indices <- sample(nrow(find_k), size = 1000)  # You can adjust the sample size as needed
sample_data <- find_k[sample_indices, ]

# Convert the sample data to a matrix
sample_matrix <- as.matrix(sample_data)

# Compute the number of clusters using the within-cluster sum of squares (wss)
wss <- fviz_nbclust(sample_matrix, kmeans, method = "wss")
gap_stat <- fviz_nbclust(sample_matrix, kmeans, method = "gap_stat")
silhouette <- fviz_nbclust(sample_matrix, kmeans, method = "silhouette")

# Print the result
print(wss)
print(gap_stat)
print(silhouette)
```





```{r}
kmeans(selected_df_scaled,iter.max = 100,centers = 4,nstart = 10)
```














Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
