---
title: "R Notebook"
output: html_notebook
---

Component 3

steps:importing data
      removing null values
      removing outliers
      scaling the data to normalize the data
      Verifying all data points in each field are of same format

```{r}
library(dplyr)
library(ggplot2)
library(readxl)
library(lubridate)
library(caret)
setwd("G:/LILTHOMA/Rise_Wpu/Mentor_Mind/Identify_customer_segments_for_online_retail_using_K-means")
customer_data <- read_excel("Online Retail.xlsx")
View(customer_data)

null_values <- colSums(is.na(customer_data))
```


```{r}
null_df <- data.frame(null_values)
print(sum(duplicated(customer_data)))
null_df%>%mutate(percentage=null_values/nrow(customer_data))

```
 #CustomerID has 24% data missing and Description has less than 1% missing values
 #5268 duplicated values are present
```{r}
clean_df <- customer_data%>%
  filter_all(all_vars(!is.na(.)))
removed_data <- customer_data%>%
  filter_all(any_vars(is.na(.)))
clean_df
```
Now we check the str of data to understand the format of each field if there is any changes required we will implement the and edit where changes are required



```{r}
#Removing Duplicates
final_df <-distinct(clean_df)
print(sum(duplicated(final_df)))
```
#We have data from december 2010 to november 2011
```{r}
final_df$month <-month(final_df$InvoiceDate)
summary(final_df$month)
```

```{r}
#finding outliers UnitPrice

q1_price <- quantile(final_df$UnitPrice,0.25,na.rm=T)
q3_price <-quantile(final_df$UnitPrice,0.75,na.rm=T)
iqr<-q3_price - q1_price
lower_bound_price = q1_price - 1.5*iqr
upper_bound_price =q3_price + 1.5*iqr
outliers <- final_df$UnitPrice[final_df$UnitPrice < lower_bound_price | final_df$UnitPrice > upper_bound_price]
final_df$UnitPrice <- ifelse(final_df$UnitPrice %in% outliers, NA, final_df$UnitPrice)


```


```{r}
boxplot(final_df$UnitPrice)
```


```{r}
#finding outliers Quantity

q1_quant <- quantile(final_df$Quantity,0.25,na.rm=T)
q3_quant <-quantile(final_df$Quantity,0.75,na.rm=T)
iqr<-q3_quant - q1_quant
lower_bound_quant = q1_quant - 1.5*iqr
upper_bound_quant =q3_quant + 1.5*iqr
outliers_quant <- final_df$Quantity[final_df$Quantity < lower_bound_quant | final_df$Quantity > upper_bound_quant]
final_df$Quantity <- ifelse(final_df$Quantity %in% outliers_quant, NA, final_df$Quantity)
boxplot(final_df$Quantity)

```


```{r}
sum(is.na(final_df))
new_data <- final_df%>%filter_all(all_vars(!is.na(.)))
sum(is.na(new_data))
```
```{r}
print(new_data%>%group_by(CustomerID)%>%count()%>%arrange(-n)%>%head(30))
new_data%>%group_by(StockCode)%>%count()%>%arrange(-n)%>%head(30)
new_data%>%group_by(InvoiceNo)%>%count()%>%arrange(-n)%>%head(30)
sum(duplicated(new_data))

```


#We have finished the data preperation process
1)removed null values
2)extracted month values
3)removed outiers from the data

#My thoughts & Further process
* invoice number stock code and customer Id are the unique fields and is not benefical for this clustering still i am going to clean the inconsistencies in these fields

*Conduct a EDA before going on towards RFM analysis to understand the trends

* I am familiar with RFM analysis a marketing technique for customer segementation so i am going to create some features for Recency Frequency and monetary and a combiation of these is RFM value

*Correct Skewness in the data 

*Conduct EDA on RFM data Answer some key questions then  to standardise and clustering

*To identify the features of customers in each of Clusters
```{r}
#library(stringr)
#new_data$StockCode <- trimws(str_replace(new_data$StockCode, "[A-Za-z]$", ""))
#new_data$StockCode <- as.numeric(new_data$StockCode)
#new_data$InvoiceNo <- trimws(str_replace(new_data$InvoiceNo, "[A-Za-z]$", ""))
#new_data$InvoiceNo <- as.numeric(new_data$InvoiceNo)
#str(new_data)
```
Above Codes has been commented out because it indroduces a lot of NA values and we are not going to use those columns for clustering 

```{r}
#Final cleaned values with no inconsistent data formats and no null or duplicate values
#Quantity column contains negative values which should be removed to actually have sensible data
cleaned_data <-new_data%>%filter_all(all_vars(!is.na(.)))
sum(is.na(cleaned_data))
cleaned_data <- cleaned_data%>%filter(Quantity > 0)
#Creating Revenue column
cleaned_data$revenue <- cleaned_data$UnitPrice * cleaned_data$Quantity 

```


#EDA
```{r}
monthly_revenue <- cleaned_data%>%group_by(month)%>%summarise(sum=sum(revenue))
ggplot(data=monthly_revenue,aes(x=month,y=sum))+
  geom_line(color="Blue")+
  labs(title="The trend of Revenue for different Months",x="months",y="Revenue")+
  theme_minimal()
monthly_orders <- cleaned_data%>%group_by(month)%>%summarise(sum=n())
ggplot(data=monthly_orders,aes(x=month,y=sum))+
  geom_line(color="green")+
  labs(title="The trend  for no of orders in  different Months",x="months",y="No_of_orders")+
  theme_minimal()
```
```{r}
#Is there any seasonal customers in our data because we can see that from the months 
#August to November there is a significant rise in orders which result in more revenue
cleaned_data%>%group_by(CustomerID,month)%>%count()

#Lets Not take month into consideration not because we can surely see a uptrend from months of august to november because they may just a seasonal trend  But we cannot also just neglect the fact that buying power of customers are most high in these months but for our purpose currently it not required
```


#Feature Engineering
```{r}
summary(cleaned_data)
#Recency I will take difference in days from last date of purchase in this dataset
#Frequency :Here the largest values are given score of 1 and d
#Monetary :here also the largest values are ranked one 
#These ranking will be based on the Quatile values
last_date <- min(cleaned_data$InvoiceDate)
recency_df <- cleaned_data%>%group_by(CustomerID)%>%summarise(Last_purchase=max(InvoiceDate))
recency_df$Recency <- round(difftime(recency_df$Last_purchase,last_date,units = "days"))
recency_df <- recency_df[,c(1,3)]
recency_df$Recency =as.numeric(recency_df$Recency)


freq_monetary_df <-cleaned_data%>%group_by(CustomerID)%>%
  summarise(monetary = sum(revenue),freq = n())

rfm_table <- inner_join(recency_df,freq_monetary_df,by="CustomerID")
summary(rfm_table)
```
# We have so far created a column for monetary frequency and Recency Now we will asssign scores for these using some basic logic and created RFM value which basically is the concatenation of
Recency Score + Frequency Score +Monetary Score 
ex : 1 + 2+ 4  = 124  # This is not addition this is concatenation

```{r}

rfm_table$Recency_score <- if_else(rfm_table$Recency < 230 ,1,
                                   if_else(rfm_table$Recency < 323 ,2,
                                           if_else(rfm_table$Recency <356 ,3,4))) 
rfm_table$Freq_score <- if_else(rfm_table$freq > 87 ,1,
                                if_else(rfm_table$freq >36,2,
                                        if_else(rfm_table$freq > 14 ,3,4)))
rfm_table$Monetary_score <- if_else(rfm_table$monetary > 1126 ,1,
                                    if_else(rfm_table$monetary >466 ,2,
                                            if_else(rfm_table$monetary > 206 ,3,4)))
#Here use sep="" inside paste function to avoid NAs during corecion 
rfm_table$rfm_value <- as.numeric(paste(rfm_table$Recency_score,rfm_table$Freq_score,rfm_table$Monetary_score,sep=""))
#View(rfm_table)


```
RFM segmentation readily shows customer for any business like Best Customers, Loyal Customer, Customers on the verge of losing, Highest revenue-generating customers etc.

```{r}
print(paste("Best Customer :" ,rfm_table%>%filter(rfm_value ==111)%>%count()))
print(paste("Loyal Customer :" ,rfm_table%>%filter(Freq_score==1)%>%count()))
print(paste("Big Spenders :" ,rfm_table%>%filter(Monetary_score==1)%>%count()))
print(paste("Risk of Loosing :" ,rfm_table%>%filter(rfm_value==134)%>%count()))
print(paste("Lost best customers :" ,rfm_table%>%filter(rfm_value==411)%>%count()))
print(paste("Lost :" ,rfm_table%>%filter(rfm_value==344)%>%count()))
print(paste("Lost Cheap Customer :" ,rfm_table%>%filter(rfm_value==444)%>%count()))
#like this we can make out a lot of categories from RFM value

```
#Before using k-means to the data
we have to make sure that

Data’s distribution is not skewed    
Data is standardised (i.e. mean of 0 and standard deviation of 1).

You can use log transformation or min max scaling of R package to normalise your data here i am using log
```{r}
library(moments)
#Visualising the distributions
feature_cols <- c("Recency","freq","monetary")

par(mfrow=c(3,1))
for (feature in feature_cols){
  hist(rfm_table[[feature]],col="blue",border="black",breaks = 30,main = paste("histogram of",feature,round(skewness(rfm_table[[feature]])),2))
}
par(mfrow=c(3,1))
for (feature in feature_cols){
  plot(density(rfm_table[[feature]]),main = paste("Density of",feature))
}

```

```{r}
# Selecting columns for clustering
names(rfm_table)
data_for_clustering <- rfm_table[,c(1:4)]

#We will only transform monetary and frequency to reduce skewness  

data_for_clustering$freq    <- log(data_for_clustering$freq+1)
data_for_clustering$monetary    <- log(data_for_clustering$monetary+1)

skewness(rfm_table)
skewness(data_for_clustering)


new_rfm_table <-data_for_clustering
```

```{r}
par(mfrow=c(3,1))
for (feature in feature_cols){
  hist(new_rfm_table[[feature]],col="blue",border="black",breaks = 30,main = paste("histogram of",feature,round(skewness(new_rfm_table[[feature]])),2))
}
par(mfrow=c(3,1))
for (feature in feature_cols){
  plot(density(new_rfm_table[[feature]]),main = paste("Density of",feature))
}
```




```{r}

# Define preprocessing steps (centering scaling)
preprocess_params <- preProcess(new_rfm_table[, -1], method = c("scale","center"))

# Apply centering and scaling to the data
normalized_scaled_data <- predict(preprocess_params, newdata = new_rfm_table )

final_data <- normalized_scaled_data
selected_df_scaled <-normalized_scaled_data

```



```{r}
#install.packages("factoextra")
library(factoextra)
#methods : wss silhouette gap_stat
wss<- fviz_nbclust(selected_df_scaled,kmeans,method="wss")
#gap_stat <- fviz_nbclust(selected_df_scaled, kmeans, method = "gap_stat")
#silhouette <- fviz_nbclust(selected_df_scaled, kmeans, method = "silhouette")

# Print the result
print(wss)
#print(gap_stat)
#print(silhouette)


```




```{r}
selected_df <- selected_df_scaled
find_k <-selected_df[,-1] 
set.seed(123)
sample_indices <- sample(nrow(find_k), size = 1000)  # You can adjust the sample size as needed
sample_data <- find_k[sample_indices, ]

# Convert the sample data to a matrix
sample_matrix <- as.matrix(sample_data)

# Compute the number of clusters using the within-cluster sum of squares (wss)
wss <- fviz_nbclust(sample_matrix, kmeans, method = "wss")
gap_stat <- fviz_nbclust(sample_matrix, kmeans, method = "gap_stat")
silhouette <- fviz_nbclust(sample_matrix, kmeans, method = "silhouette")

# Print the result
print(wss)
print(gap_stat)
print(silhouette)
```





```{r}
km3 <- kmeans(selected_df_scaled[,-1],iter.max = 100,centers = 3,nstart = 10)
km4 <- kmeans(selected_df_scaled[,-1],iter.max = 100,centers = 4,nstart = 10)
km5 <- kmeans(selected_df_scaled[,-1],iter.max = 100,centers = 5,nstart = 10)
km3
km4
km5
```


```{r}
fviz_cluster(list(data = selected_df_scaled,cluster=km3$cluster))
fviz_cluster(list(data = selected_df_scaled,cluster=km4$cluster))
fviz_cluster(list(data = selected_df_scaled,cluster=km5$cluster))


```



```{r}
og_data <- rfm_table
og_data$cluster <- km4$cluster
og_data%>%group_by(cluster)%>%
  summarise(r_mean = mean(Recency,trim=1),f_mean=mean(freq),m_mean=mean(monetary),cnt=n())

og_scaled <- selected_df_scaled
og_scaled$cluster <- km4$cluster
og_scaled%>%group_by(cluster)%>%
  summarise(r_mean = mean(Recency,trim=1),f_mean=mean(freq),m_mean=mean(monetary),cnt=n())

library(reshape2)
selected_df$cluster <-km4$cluster
selected_df
df_melt <-melt(data=selected_df[,-1],id.vars ="cluster",variable.name = "metric",value.name = "value" )


ggplot(df_melt, aes(x = metric, y = value, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~cluster, nrow = 1) +
  theme_minimal()


```



```{r}
library(dplyr)

# Calculate average values for each metric within each cluster
avg_values <- df_melt %>%
  group_by(cluster, metric) %>%
  summarise(avg_value = mean(value)) %>%
  ungroup()


# Create the snake plot
ggplot(avg_values, aes(x = metric, y = avg_value, color = as.factor(cluster))) +
  geom_line(aes(group = cluster), size = 1.5) +
  geom_point(size = 3) +
  labs(title = "Snake Plot", x = "Metric", y = "Average Value") +
  theme_minimal()

```

Cluster 1 : Best Customer                                Platinum
Cluster 2 : Loyal Customers                              Gold
Cluster 3 : Loosing Customers                            Bronze
Cluster 4 : Passing by or Newer Customers                Silver




Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
